---
title: "Filter JABOT specimen records"
format: html
execute:
  echo: false
  eval: true
  freeze: auto
  cache: false
  warning: false
  message: false
  include: true
  inline: true
---

```{r setup, include=FALSE}
library(jabotR)
```

This guide demonstrates how to use the `jabot_records()` function in the `jabotR` package to retrieve specimen occurrence records for specific taxa from the Rio de Janeiro Botanical Garden's JABOT online managing herbarium collection system.

## Function Overview

The `jabot_records()` function retrieves filtered plant specimen records from JABOT collections based on taxon, herbarium, location, and date criteria. It handles downloading, parsing, filtering, and optionally saving the occurrence data.

This tool is ideal for floristic, taxonomic, or biogeographic research, especially when working with large datasets for a particular plant group.

## Arguments

| Argument      | Description |
|---------------|-------------|
| `herbarium`   | Vector of herbarium codes (e.g., `"RB"`, `"SP"`). Use `NULL` for all collections. |
| `repatriated` | Logical. If `FALSE`, skips repatriated herbaria. |
| `taxon`       | One or multiple family, genus, or species names to filter by. |
| `state`       | Vector of Brazilian states (e.g., `"BA"`, `"SP"`). |
| `recordYear`  | A specific year (e.g., `"2001"`) or range (e.g., `c("1990", "2024")`). |
| `indets`      | If `FALSE`, filters out records not identified to species level. |
| `reorder`     | Controls column order of the final result (e.g., by herbarium, taxon, year). |
| `path`        | Optional directory containing existing JABOT DwC-A files. |
| `updates`     | If `TRUE`, checks for updated versions of data in the IPT. |
| `verbose`     | If `TRUE`, shows progress messages. |
| `save`        | If `TRUE`, saves results as CSV and log files. |
| `dir`         | Folder where output will be saved. |
| `filename`    | Name of the resulting CSV and log files. |

## Basic Use Case

```r
jabot_records(
  taxon = c("Fabaceae", "Ochnaceae"),
  verbose = TRUE,
  save = TRUE,
  dir = "jabot_records",
  filename = "fabaceae_ochnaceae_records"
)
```

## Filter by Herbarium and Year Range

```r
jabot_records(
  taxon = "Fabaceae",
  herbarium = c("R", "RB"),
  recordYear = c("1995", "2023"),
  verbose = TRUE,
  save = FALSE
)
```

## Filter by State and Remove Indeterminate Records

```r
jabot_records(
  taxon = "Malpighiaceae",
  state = c("BA", "PE"),
  indets = FALSE,
  save = TRUE,
  dir = "malpighiaceae_ba_pe"
)
```

## Using a Pre-Downloaded Dataset

```r
jabot_records(
  path = "jabot_download",
  taxon = "Myrtaceae",
  updates = FALSE,
  save = FALSE
)
```

## Visualizing Specimens by Year

```r
library(dplyr)
library(ggplot2)

records <- jabot_records(taxon = "Fabaceae", 
                          herbarium = "RB", 
                          recordYear = c("2000", "2025"), 
                          save = FALSE)

records %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Fabaceae Records by Year (RB Herbarium)",
       x = "",
       y = "Number of Records")
```

```{r}
#| echo: false
library(dplyr)
library(ggplot2)

records <- jabot_records(taxon = "Fabaceae", 
                         herbarium = "RB", 
                         recordYear = c("2000", "2025"),
                         verbose = FALSE,
                         save = FALSE)

records %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "#81ae7a") +
  labs(title = "Fabaceae Records by Year (RB Herbarium)",
       x = "",
       y = "Number of Records") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

## Tips

- Use `recordYear = c("YYYY", "YYYY")` to analyze records over a time span.
- Use `jabot_summary()` beforehand to check herbarium availability.
- Set `save = TRUE` for reproducible outputs and logs.

## See Also

- [`jabot_download()`](/articles/download_jabot_collections.qmd): Download JABOT specimen records
- [`jabot_parse()`](/articles/parse_jabot_dwca.qmd): Parse JABOT archive files
- [`jabot_summary()`](/articles/summarize_jabot_collections.qmd): Summarize JABOT collections
- [`jabot_indets()`](/articles/filter_indeterminate_jabot_specimens.qmd): Extract indeterminate records
