---
title: "Organize field data with the jabotR field notebook"
format: html
execute:
  echo: true
  eval: false
  freeze: auto
  cache: false
  warning: false
  message: false
  include: true
  inline: true
---

```{r setup, include=FALSE}
library(jabotR)
```

Here we present a complete guide to `jabot_fieldbook()`, the interactive field notebook distributed with the `jabotR` package. Unlike the other `jabotR` functions, which query the JABOT web service, the field notebook is a self-contained tool for organizing new collection data in the field and turning it into a spreadsheet ready for submission to JABOT.

## Function Overview

`jabot_fieldbook()` opens the field notebook, built with plain HTML, CSS and JavaScript, in the user's default web browser. It runs entirely offline, requires neither Shiny nor an internet connection, and performs all spreadsheet generation locally in the browser — no data ever leaves the user's computer. The interface is available in Portuguese (the default) and English, with a language switch in the bottom-left corner.

## Arguments

| Argument  | Description |
|-----------|-------------|
| `browser` | Logical. If `TRUE` (default), opens the field notebook in the default web browser. If `FALSE`, the application is not opened and only its local path is returned. |

## Opening the Field Notebook

```{r}
jabot_fieldbook()
```

To retrieve the local path to the application (for example, to open it manually or embed it elsewhere) without launching a browser tab:

```{r}
app_path <- jabot_fieldbook(browser = FALSE)
app_path
```

## How Data Are Organized

The field notebook is built around **collection events**. A collection event represents a single collecting date/locality combination — the equivalent of one field-notebook entry — and groups together:

- One shared **header**, entered once and reused by every specimen collected during that event (date, country, state, municipality, locality, protected area, habitat, coordinates, altitude, and collector information).
- One or more **specimens**, each with its own taxonomic and specimen-level data (collection number, family, genus, species, determiner, plant description, and other fields).

A running summary at the top of the page reports, in real time, the number of collection events, the number of specimens, the percentage of headers that are complete, and the fixed number of columns (58) in the JABOT spreadsheet format that the tool exports to.

Click **Add collection event** to start a new event, or use **Repeat previous header** / **Duplicate header** inside an existing event to reuse locality and date information across nearby collecting events without retyping it.

### Header fields

| Field | Description |
|---|---|
| Day / Month / Year | Collection date. |
| Country, State, Municipality | Political divisions where the collection took place. |
| Locality | Locality within the municipality (farm, river, road, community, etc.). |
| Protected area (UC / TI) | Conservation unit, Indigenous Territory, or other protected area. |
| Habitat | Description of the specimen's habitat. |
| Latitude / Longitude | Coordinates in decimal degrees. |
| Altitude | Minimum altitude or depth. |
| Collector / Additional collectors | Main collector and any additional collectors. |
| Locality notes | Free-text notes about the collecting locality. |

An **Advanced header fields** panel exposes further, less frequently needed columns: coordinates split into degrees/minutes/seconds with N/S and E/W hemisphere, maximum altitude and its unit, the source herbarium acronym, duplicate herbaria and their count, and the associated project.

### Specimen fields

Each row of the specimen table covers the fields needed for a quick first pass — collection number, family, genus, `cf.` qualifier, species epithet, determiner, determination date, and a free-text plant description. Column widths can be adjusted directly in the table.

Opening **Details** on a specimen reveals the complete record, organized into four groups:

| Group | Fields |
|---|---|
| Taxonomic identification | Family, genus, `cf.`, species 1–3 with authors and infraspecific ranks, vernacular name, type status |
| Collection and determination | Number, determiner, determination date, accession number (`numtombo`), suffix |
| Individual description | Flowers, fruits, stem, height and its unit, habit |
| Uses and notes | Uses, specific use, plant description |

Use **Add specimen** to add a new row, **Duplicate last** to copy the previous specimen as a starting point, or **Continue number sequence** to auto-increment the collection number for a new specimen.

## Reviewing Before Export

A validation panel checks the data as it is entered and flags, per collection event, issues such as an incomplete collection date, a missing collector, an event with no specimens, or a specimen missing its collection number or any taxonomic information (at least family, genus, or species). Review the flagged items before exporting; the spreadsheet can still be generated with warnings pending if needed.

## Saving and Loading Drafts

- **Save draft** stores the current state of the notebook in the browser's local storage, so work in progress is not lost if the tab is closed.
- **Load draft** restores the most recently saved draft.

Drafts are kept locally in the browser being used and are not synced across devices or browsers — save a draft again after significant changes, and avoid clearing browser site data before exporting.

## Exporting

Three export options are available, at the top and bottom of the page:

- **Download empty spreadsheet**: generates a blank spreadsheet with the 58 JABOT columns and no data, useful as a template or for entering data directly in a spreadsheet application.
- **Generate Jabot spreadsheet**: exports all completed collection events and specimens as a single spreadsheet in the standard JABOT column layout, ready for submission.
- **Generate field notebook (PDF)**: renders a printable field notebook from the entered data, useful as a physical or archival record of the collecting trip.

## Additional Tips

- Fill in the header of a collection event first, then add its specimens — shared header fields are copied automatically to every specimen in that event when the spreadsheet is generated.
- Use **Repeat previous header** when starting a new event at the same locality and date as the previous one (e.g., several specimens collected together but logged as separate events).
- Save a draft periodically, particularly during long field sessions, since the field notebook keeps no server-side copy of the data.
- The field notebook works fully offline, so it can be used directly in the field on a laptop with no signal; only exporting figures/spreadsheets requires no additional steps once back online.

## See Also

- [`jabot_records()`](/articles/filter_jabot_records.qmd): Retrieve and filter existing JABOT specimen records
- [`jabot_summary()`](/articles/summarize_jabot_collections.qmd): Get a summary of JABOT collections
